A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network

### 2018-11-01 anzjs2018/anzjs2018.tex
edits
run spellcheck

\doublespace

@@

\begin{abstract}
Predicting the arrival time of a transit vehicle involves not only
knowledge of its current position and schedule adherance,
but also of traffic conditions along the remainder of the route.
Road networks are dynamic, 
and can change from free-flowing to highly congested very quickly,
which impacts the arrival time of transit vehicles,
particularly buses which often share the road with other vehicles,
so reliable predictions need to account for \rt and future traffic conditions.
The first step in this process is to construct a framework with which road state
(traffic conditions) can be estimated using \rt transit vehicle position data.
We propose a framework which implements a vehicle model using a particle filter
to estimate travel times along roads,
used by a second model to estimate \rt traffic conditions which can
later be used for arrival time prediction.
We demonstrate the \rt faesibility of such an approach in \rt,
and detail our implementation in R and C++.
Development and testing took place in Auckland, New Zealand,
with future work planned to incorporate historical data into the network model
to improve the reliability of ETAs.
\end{abstract}

>>>>

% \doublespace
\usepackage{afterpage}

@@

\begin{abstract}
Predicting the arrival time of a transit vehicle involves not only
knowledge of its current position and schedule adherence,
but also of traffic conditions along the remainder of the route.
Road networks are dynamic, 
and can change from free-flowing to highly congested very quickly,
which impacts the arrival time of transit vehicles,
particularly buses which often share the road with other vehicles,
so reliable predictions need to account for \rt and future traffic conditions.
The first step in this process is to construct a framework with which road state
(traffic conditions) can be estimated using \rt transit vehicle position data.
We propose a framework which implements a vehicle model using a particle filter
to estimate travel times along roads,
used by a second model to estimate \rt traffic conditions which can
later be used for arrival time prediction.
We demonstrate the \rt feasibility of such an approach in \rt,
and detail our implementation in R and C++.
Development and testing took place in Auckland, New Zealand,
with future work planned to incorporate historical data into the network model
to improve the reliability of ETAs.
\end{abstract}

### 2018-11-01 anzjs2018/sections/01_introduction.tex
run spellcheck

To avoid the problems discussed above,
a more robust modeling and prediction framework is required.
This should consist of a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data;
and a way to incorporate road congestion information 
into the arrival time predictions.
Several vehicle modeling approaches were explored, 
including the Kalman filter \citep{Dailey_2001,Cathey_2003},
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter \citep{Hans_2015}.
The last of these was chosen for several reasons
which are discussed in Section~\ref{sec:pf}.

The next step is to construct a method of estimating traffic conditions.
This can be done by using the travel time of preceeding buses along the road
to predict travel times for the current bus.
\cite{Yu_2010} obtained improved prediction accuracy by using travel times
from preceeding buses servicing the same route.
\cite{Hans_2015} used a similar approach,
instead referring to headway, the time between consecutive vehicles
at a point on the route.
However, this approach only works well for high frequency routes.
\cite{Yu_2011} showed further improvements by combining travel times 
from several routes which converged,
but is limited to predefined, converging routes.
In general, however, no comprehensive network modeling approach has been proposed using
solely GPS position data to accounting for congestion when estimating arrival times.
We describe a simple method of constructing a road network
from GTFS data (section~\ref{sec:gtfs}),
which allows all transit vehicles traveling along a road to contribute to tis state,
regardless of which route they are serving.

In this paper, we describe a \rt vehicle and network state
model which uses the vehicles traveling throughout the transit road network
to update congestion information in real-time.
As the vehicle model is implemented using a particle filter,
a computationally intensive method,
we demonstrate its faesibility as a valid real-time option.
The vehicle model provides estimates of travel times along roads,
which can be used to update the road network's state.
Future work is planned to use this travel time information 
in combination with historical data to develop an arrival time prediction model.

>>>>

To avoid the problems discussed above,
a more robust modelling and prediction framework is required.
This should consist of a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data;
and a way to incorporate road congestion information 
into the arrival time predictions.
Several vehicle modelling approaches were explored, 
including the Kalman filter \citep{Dailey_2001,Cathey_2003},
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter \citep{Hans_2015}.
The last of these was chosen for several reasons
which are discussed in Section~\ref{sec:pf}.

The next step is to construct a method of estimating traffic conditions.
This can be done by using the travel time of preceding buses along the road
to predict travel times for the current bus.
\cite{Yu_2010} obtained improved prediction accuracy by using travel times
from preceding buses servicing the same route.
\cite{Hans_2015} used a similar approach,
instead referring to headway, the time between consecutive vehicles
at a point on the route.
However, this approach only works well for high frequency routes.
\cite{Yu_2011} showed further improvements by combining travel times 
from several routes which converged,
but is limited to predefined, converging routes.
In general, however, no comprehensive network modelling approach has been proposed using
solely GPS position data to accounting for congestion when estimating arrival times.
We describe a simple method of constructing a road network
from GTFS data (section~\ref{sec:gtfs}),
which allows all transit vehicles travelling along a road to contribute to its state,
regardless of which route they are serving.

In this paper, we describe a \rt vehicle and network state
model which uses the vehicles travelling throughout the transit road network
to update congestion information in real-time.
As the vehicle model is implemented using a particle filter,
a computationally intensive method,
we demonstrate its feasibility as a valid real-time option.
The vehicle model provides estimates of travel times along roads,
which can be used to update the road network's state.
Future work is planned to use this travel time information 
in combination with historical data to develop an arrival time prediction model.

### 2018-11-01 anzjs2018/sections/02_gtfs.tex
run spellcheck

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about
\begin{itemize}
\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
\item \emph{routes}, a sequence of two or more stops displayed as a single service;
\item \emph{trips}, an instance of a route occuring at a specific time of day;
\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
\item \emph{shapes}, the sequence of points defining a vehicle's path along a route
\end{itemize}
\citep{GoogleDevelopers_2006}.
The second component is \emph{GTFS realtime},
which is only available in a subset of the providers due to the requirement of 
onboard GPS tracking devices and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
and are typically accessed by developers via an API can be used in \rt applications.

@@

The primary predictor of arrival time is 
the travel time along roads between where the bus is \emph{now},
and the stop at which the passenger is waiting.
In most applications, however, this vital information is unavailable,
at least directly,
so here we construct a transit road network consisting of intersections
and the roads between them.
By generalising to physical roads,
we remove any route-dependency,
which increases the quantity of data available for each road.
Each route is then represeted as a sequence of road segments,
each with an associated travel time.

The transit network is constructed by splitting each route
into spatially identifiable segments,
each representative of a physical road.
The simplest way to do this is to use bus stops as nodes in the network
(Figure~\ref{fig:network_creation_1}),
and connecting roads as edges.
In this way, routes that serice the same subsequence of stops
will all contribute to congestion information for the connecting roads,
as shown in Figure~\ref{fig:network_creation_2}.
There are several drawbacks to this method,
for example routes merge at physical intersections, not bus stops,
so some overlap will occur as shown.
However, this is minimal, and our simplification is sufficent for the current work.

@@

\subsection{Realtime vehicle locations}
\label{sec:realtime_data}

@@

One important consideration regarding Auckland Transport's \rt implementation is that
buses are programmed to report their location when arriving at or departing from
bus stops, as well as some major intersections.
To further complicate this,
these positions can be pre-emtive,
(i.e., the bus is almost there, but not quite),
so consecutive observations can show a bus traveling backwards.
To handle this, we compute the approximate distance traveled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the current state is rejected and the vehicle reverted
to its previous state before continuing.
This way, we avoid degeneration of the particle fitler (Section~\ref{sec:model_perf})
at the cost of losing one observation that is most likely invalid anyway.

>>>>

There are two components to GTFS.
The first, \emph{GTFS static}, includes information about
\begin{itemize}
\item \emph{stops}, a physical location where passengers can embark and disembark the vehicle;
\item \emph{routes}, a sequence of two or more stops displayed as a single service;
\item \emph{trips}, an instance of a route occurring at a specific time of day;
\item \emph{schedules}, specifying the arrival (and departure) times for each bus at each of its stops; 
\item \emph{shapes}, the sequence of points defining a vehicle's path along a route
\end{itemize}
\citep{GoogleDevelopers_2006}.
The second component is \emph{GTFS realtime},
which is only available in a subset of the providers due to the requirement of 
on board GPS tracking devices and a central server.
It provides a standardised format for sharing vehicle positions and trip delays,
and are typically accessed by developers via an API can be used in \rt applications.

@@

The primary predictor of arrival time is 
the travel time along roads between where the bus is \emph{now},
and the stop at which the passenger is waiting.
In most applications, however, this vital information is unavailable,
at least directly,
so here we construct a transit road network consisting of intersections
and the roads between them.
By generalising to physical roads,
we remove any route-dependency,
which increases the quantity of data available for each road.
Each route is then represented as a sequence of road segments,
each with an associated travel time.

The transit network is constructed by splitting each route
into spatially identifiable segments,
each representative of a physical road.
The simplest way to do this is to use bus stops as nodes in the network
(Figure~\ref{fig:network_creation_1}),
and connecting roads as edges.
In this way, routes that service the same subsequence of stops
will all contribute to congestion information for the connecting roads,
as shown in Figure~\ref{fig:network_creation_2}.
There are several drawbacks to this method,
for example routes merge at physical intersections, not bus stops,
so some overlap will occur as shown.
However, this is minimal, and our simplification is sufficient for the current work.

@@

\subsection{Real-time vehicle locations}
\label{sec:realtime_data}

@@

One important consideration regarding Auckland Transport's \rt implementation is that
buses are programmed to report their location when arriving at or departing from
bus stops, as well as some major intersections.
To further complicate this,
these positions can be preemptive,
(i.e., the bus is almost there, but not quite),
so consecutive observations can show a bus travelling backwards.
To handle this, we compute the approximate distance travelled, $\tilde x_k$,
of the vehicle by finding the nearest point on the path to the observation;
if this has decreased, the current state is rejected and the vehicle reverted
to its previous state before continuing.
This way, we avoid degeneration of the particle filter (Section~\ref{sec:model_perf})
at the cost of losing one observation that is most likely invalid anyway.

### 2018-11-01 anzjs2018/sections/03_models.tex
run spellcheck

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance traveled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
Additionally, travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
are included in the state and estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector,
\begin{equation}
\label{eq:vehicle_state}
\bX_k = 
\begin{bmatrix}
    x_k \\ \dot x_k \\ \bz
\end{bmatrix}.
\end{equation}
Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
with GPS error $\epsilon^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
\begin{equation}
\label{eq:vehicle_obs}
\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
\end{equation}

We use a particle filter to estimate $\bX_k$,
due to do its high flexibility and use in recent 
transit vehicle modeling applications \citep{Hans_2015}.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is the intuitive likelihood function, 
discussed in section~\ref{sec:pf_update}.
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

\begin{figure}[p]
    \centering
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_1.pdf}
        \caption{
            Vehicle state is mapped to observation space by the
            measurement function $h$.   
        }
        % The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
        \label{fig:pf_state_prev}
    \end{subfigure}\\
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_2.pdf}
        \caption{
            The transition function $f$ predicts the future state 
            of each particle.
        }
            % The transition function $f$ models the behaviour of each particle,
            % including changes in acceleration and stopping behaviour at bus stops,
            % which can result in multiple modes.}
        \label{fig:pf_state_predict}
    \end{subfigure}\\
    % \begin{subfigure}[t]{0.48\textwidth}
    %     \centering
    %     \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
    %     \caption{Each particle's distance from the true observation (red cross) 
    %         is calculated using
    %         the geographic distance between two coordinates, which can then be used
    %         to calculate it's weight.}
    %     \label{fig:pf_state_update}
    % \end{subfigure}\;\;
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_4.pdf}
        \caption{
            The state is updated by weighted resampling based on distance
            of particle to observation.
        }
        % Resampling occurs using likelihood weights that use the distance
        %     between the particle's location and the vehicles reported location (red cross),
        %     resulting in the updated state.}
        \label{fig:pf_state_predict2}
    \end{subfigure}
    \caption{
        The particle filter approximates vehicle state using a set of 
        discrete points (a), which are each mutated indepently (b)
        to predict the next state.
        The measurement function $h$ allows state particle state
        to be mapped to a GPS coordinate (a),
        which can be compared to the observed location (c)
        to calculate each particles likelihood and perform resampling.
    }
        % The vehicles state is estimated using a set of discrete points, or particles,
        % each of which is independently mutated to make predictions of the future state.
        % Multimodality is easily handled, as demonstrated here,
        % without degeneration of the filter.}
    \label{fig:pf_state}
\end{figure}

@@

In a particle filter, the posterior distribution of the state at time $t_{k-1}$
is represented by a set of discrete points, or particles, (Figure~\ref{fig:pf_state_prev}),
each with an associated weight $W_{k-1}^{(i)}$
\begin{equation}
p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
:= \left\{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation}
each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
\begin{equation}
p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
\left\{f(\bX_{k-1}^{(i)}, \psi), W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation}
using parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, 
the state is updated by reweighting the particles using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
(sec.~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
\begin{equation*}
W_k^{(i)} = \frac{W_{k-1}^{(i)} p(\bY_k | \bX_k^{(i)})}{
    \sum_{j=1}^N W_{k-1}^{(j)} p(\bY_k | \bX_k^{(j)})
}
\end{equation*}
generating a posterior estimate of the state
\begin{equation}
p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
\left\{\bX_{k}^{(i)}, W_{k}^{(i)}\right\}_{i=1}^N.
\end{equation}

In order to prevent degeneration of the particle filter state,
\emph{importance resampling} is employed whenever the 
effective sample size 
\begin{equation}
\label{eq:neff}
N_{\text{eff}} = \frac{1}{\sum_i (W_k^{(i)})^2}
\end{equation}
falls below a specified threshold $N_{\text{thres}}$
\citep{}.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_predict2}.
Afterwards, the weights are reinitialized to $W_k^{(i)} = N^{-1}$.
Note this is a computationally demanding step,
so reducing the rate of resampling is favourable.

@@

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's speed changes as a Markov process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way,
so the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close,
and account for deceleration and acceleration.
For these parameters, we used constant values for all stops,
based on the values used by \cite{Hans_2015};
future work will investigate modeling these in real-time similarly to Section~\ref{sec:kf}.

The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
which implements those features discussed above.
The stop positions are represented by their distance into trip,
$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$.
The $stop\_index$ function returns the index of a stop,
given a distance traveled and set of stop distances.
Lines~2--9 are necessary to decrease the degeneration rate of the filter,
which can occur when there is a long delay between observations
simply because the bus has not moved due to traffic or other reasons. 
We therefore place positive probability for some particles to remain stationary.

@@

\renewcommand{\algorithmicrequire}{\textbf{Start:}}
\newcommand{\algorithmicbreak}{\textbf{break}}
\newcommand{\Break}{\State \algorithmicbreak}
\begin{algorithm}[tb]
    \caption{Particle mutation function.}
    \label{fig:algorithm}
    \begin{algorithmic}[2]
    \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
    Initialization $\delta \gets t_k - t_{k-1}$, 
        $j \gets stop\_index(x, \boldsymbol{S})$
    \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
        \Comment{Remain longer at stop}
        \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
        \State $\delta \gets \delta - \gamma - d_j$
    \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
        \Comment{Occasionally bus stopped due to traffic}
        \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
        \State $\delta \gets \delta - w$
        \State $z_j \gets z_j + w$
        \Comment{Congestion counts as travel time}
    \EndIf
    \While {$\delta > 0$ and $x < S_J$} 
        \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
        \Comment{System noise, trunctated to be non-negative}
        \State $x \leftarrow x + \dot x$
        \If {$x \geq S_{j+1}$}
            \If {$j+1 = J$}
                \Comment{Break at end of route}
                \State $x \gets S_J$
                \Break
            \EndIf
            \State $u_j \sim \mathrm{U}(0,1)$
            \Comment{Simulate vehicle stopping}
            \If {$u_j < \pi$}
                \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
                \State $\delta \gets \delta - \gamma - d_j$
                \State $x \gets S_j$
            \EndIf
            \State $j\gets j+1$
        \EndIf
        \State $\delta \gets \delta - 1$
        \State $z_j \gets z_j + 1$
        \Comment{Track travel time along road}
    \EndWhile
    \\
    \Return $x, \dot x, \boldsymbol z$
    \end{algorithmic}
\end{algorithm}

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is obtained by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection, \cite{Snyder_1998}),
allow the placement of a bivate normal likelihood on the data.
The model for the observations is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
The geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
is calculated using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible due to the discretization of the state provided by the particle filter;
in other approaches, such as the Kalman filter,
a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance traveled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.

>>>>

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance travelled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
Additionally, travel times along road segments along the route, 
$\bz = (z_1,\ldots,z_L)^\top$, in seconds,
are included in the state and estimated
sequentially as the vehicle traverses the route.
These are combined into the state vector,
\begin{equation}
\label{eq:vehicle_state}
\bX_k = 
\begin{bmatrix}
    x_k \\ \dot x_k \\ \bz
\end{bmatrix}.
\end{equation}
Observations $\bY_k$ of the vehicle are made using a GPS at time $t_k$
with GPS error $\epsilon^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle,
\begin{equation}
\label{eq:vehicle_obs}
\bY_k = \begin{bmatrix} \lambda_k \\ \phi_k \end{bmatrix}.
\end{equation}

We use a particle filter to estimate $\bX_k$,
due to do its high flexibility and use in recent 
transit vehicle modelling applications \citep{Hans_2015}.
The primary advantage of a particle filter is its handling of multimodality,
as demonstrated in figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is the intuitive likelihood function, 
discussed in section~\ref{sec:pf_update}.
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as the model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter in real-time 
and its timings.

\begin{figure}[p]
    \centering
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_1.pdf}
        \caption{
            Vehicle state is mapped to observation space by the
            measurement function $h$.   
        }
        % The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
        \label{fig:pf_state_prev}
    \end{subfigure}\\
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_2.pdf}
        \caption{
            The transition function $f$ predicts the future state 
            of each particle.
        }
            % The transition function $f$ models the behaviour of each particle,
            % including changes in acceleration and stopping behaviour at bus stops,
            % which can result in multiple modes.}
        \label{fig:pf_state_predict}
    \end{subfigure}\\
    % \begin{subfigure}[t]{0.48\textwidth}
    %     \centering
    %     \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
    %     \caption{Each particle's distance from the true observation (red cross) 
    %         is calculated using
    %         the geographic distance between two coordinates, which can then be used
    %         to calculate it's weight.}
    %     \label{fig:pf_state_update}
    % \end{subfigure}\;\;
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_4.pdf}
        \caption{
            The state is updated by weighted resampling based on distance
            of particle to observation.
        }
        % Resampling occurs using likelihood weights that use the distance
        %     between the particle's location and the vehicles reported location (red cross),
        %     resulting in the updated state.}
        \label{fig:pf_state_predict2}
    \end{subfigure}
    \caption{
        The particle filter approximates vehicle state using a set of 
        discrete points (a), which are each mutated independently (b)
        to predict the next state.
        The measurement function $h$ allows state particle state
        to be mapped to a GPS coordinate (a),
        which can be compared to the observed location (c)
        to calculate each particles likelihood and perform resampling.
    }
        % The vehicles state is estimated using a set of discrete points, or particles,
        % each of which is independently mutated to make predictions of the future state.
        % Multimodality is easily handled, as demonstrated here,
        % without degeneration of the filter.}
    \label{fig:pf_state}
\end{figure}

@@

In a particle filter, the posterior distribution of the state at time $t_{k-1}$
is represented by a set of discrete points, or particles, (Figure~\ref{fig:pf_state_prev}),
each with an associated weight $W_{k-1}^{(i)}$
\begin{equation}
p(\bX_{k-1} | \bY_{k-1}) \approx \tilde\bX_{k-1|k-1} 
:= \left\{\bX_{k-1}^{(i)}, W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation}
each of which is independently updated or \emph{mutated} using the transition function $f$ (fig~\ref{fig:pf_state_predict})
\begin{equation}
p(\bX_k | \bX_{k-1}) \approx \tilde\bX_{k|k-1} := 
\left\{f(\bX_{k-1}^{(i)}, \psi), W_{k-1}^{(i)}\right\}_{i=1}^N
\end{equation}
using parameter vector $\psi$ containing all of the necessary parameters
for the model (including system noise).
After mutation, 
the state is updated by reweighting the particles using the likelihood $p(\bY_k | \bX_k^{(i)})$ 
(Section~\ref{sec:pf_update}) and standardising so $\sum_{i=1}^N W_k^{(i)} = 1$,
\begin{equation*}
W_k^{(i)} = \frac{W_{k-1}^{(i)} p(\bY_k | \bX_k^{(i)})}{
    \sum_{j=1}^N W_{k-1}^{(j)} p(\bY_k | \bX_k^{(j)})
}
\end{equation*}
generating a posterior estimate of the state
\begin{equation}
p(\bX_k | \bY_k) \approx \tilde\bX_{k|k} := 
\left\{\bX_{k}^{(i)}, W_{k}^{(i)}\right\}_{i=1}^N.
\end{equation}

In order to prevent degeneration of the particle filter state,
\emph{importance resampling} is employed whenever the 
effective sample size 
\begin{equation}
\label{eq:neff}
N_{\text{eff}} = \frac{1}{\sum_i (W_k^{(i)})^2}
\end{equation}
falls below a specified threshold $N_{\text{thres}}$
\citep{}.
The resampling is performed with replacement using importance weights $W_k^{(i)}$,
as demonstrated in Figure~\ref{fig:pf_state_predict2}.
Afterwards, the weights are reinitialised to $W_k^{(i)} = N^{-1}$.
Note this is a computationally demanding step,
so reducing the rate of resampling is favourable.

@@

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's speed changes as a Markov process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way,
so the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close,
and account for deceleration and acceleration.
For these parameters, we used constant values for all stops,
based on the values used by \cite{Hans_2015};
future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.

The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
which implements those features discussed above.
The stop positions are represented by their distance into trip,
$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$.
The $stop\_index$ function returns the index of a stop,
given a distance travelled and set of stop distances.
Lines~2--9 are necessary to decrease the degeneration rate of the filter,
which can occur when there is a long delay between observations
simply because the bus has not moved due to traffic or other reasons. 
We therefore place positive probability for some particles to remain stationary.

@@

\renewcommand{\algorithmicrequire}{\textbf{Start:}}
\newcommand{\algorithmicbreak}{\textbf{break}}
\newcommand{\Break}{\State \algorithmicbreak}
\begin{algorithm}[tb]
    \caption{Particle mutation function.}
    \label{fig:algorithm}
    \begin{algorithmic}[2]
    \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
    Initialisation $\delta \gets t_k - t_{k-1}$, 
        $j \gets stop\_index(x, \boldsymbol{S})$
    \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
        \Comment{Remain longer at stop}
        \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
        \State $\delta \gets \delta - \gamma - d_j$
    \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
        \Comment{Occasionally bus stopped due to traffic}
        \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
        \State $\delta \gets \delta - w$
        \State $z_j \gets z_j + w$
        \Comment{Congestion counts as travel time}
    \EndIf
    \While {$\delta > 0$ and $x < S_J$} 
        \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
        \Comment{System noise, truncated to be non-negative}
        \State $x \leftarrow x + \dot x$
        \If {$x \geq S_{j+1}$}
            \If {$j+1 = J$}
                \Comment{Break at end of route}
                \State $x \gets S_J$
                \Break
            \EndIf
            \State $u_j \sim \mathrm{U}(0,1)$
            \Comment{Simulate vehicle stopping}
            \If {$u_j < \pi$}
                \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
                \State $\delta \gets \delta - \gamma - d_j$
                \State $x \gets S_j$
            \EndIf
            \State $j\gets j+1$
        \EndIf
        \State $\delta \gets \delta - 1$
        \State $z_j \gets z_j + 1$
        \Comment{Track travel time along road}
    \EndWhile
    \\
    \Return $x, \dot x, \boldsymbol z$
    \end{algorithmic}
\end{algorithm}

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is obtained by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection, \cite{Snyder_1998}),
allow the placement of a bivariate normal likelihood on the data.
The model for the observations is,
assuming GPS error $\epsilon^2$,
\begin{equation}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation}
The geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging (\ref{eq:obs_dist}) yields
\begin{equation}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right)
\end{equation}

The likelihood of the data given a particle's state estimate 
is calculated using (\ref{eq:obs_exp})
\begin{equation}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation}
It is worth noting that this representation of the likelihood is only
possible due to the discrete nature of the particle filter state estimate;
in other approaches, such as the Kalman filter,
a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance travelled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.
